---
title: "Time Serise Graphic and Decomposition"
author: "Subhalaxmi Rout"
date: "06/14/2021"
output:
  word_document: default
  html_document: default
subtitle: Week1 homework
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
library(fpp2)
library(tsibble)
```

### Forecasting: Principles and Practices (Chapter 2 - Time serise graphics)

#### 2.1

Use the help function to explore what the series gold, woolyrnq and gas represent.

**$(a)$ Use autoplot() to plot each of these in separate plots.**

```{r}
?gold
?woolyrnq
?gas
```

```{r}
autoplot(gold)
autoplot(woolyrnq)
autoplot(gas)
```

**$(b)$ What is the frequency of each series? Hint: apply the frequency() function.**

```{r}
paste0("Frequency of gold serise: ", frequency(gold)) 
paste0("Frequency of woolyrnq serise: ", frequency(woolyrnq))
paste0("Frequency of gas serise: ", frequency(gas))
```


**$(c)$ Use which.max() to spot the outlier in the gold series. Which observation was it?**

```{r}
which.max(gold)
```

The observation is 770. 

#### 2.3

Download some monthly Australian retail data from the book website. These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file.

**$(a)$ You can read the data into R with the following script:**

```{r}
retaildata <- readxl::read_excel("retail.xlsx", skip=1)
head(names(retaildata))
```

The second argument (skip=1) is required because the Excel sheet has two header rows.

**$(b)$ Select one of the time series as follows (but replace the column name with your own chosen column):**

```{r}
myts <- ts(retaildata[,"A3349627V"],
  frequency=12, start=c(1982,4))

```

**$(c)$ Explore your chosen retail time series using the following functions:**

**autoplot(), ggseasonplot(), ggsubseriesplot(), gglagplot(), ggAcf()**

**Can you spot any seasonality, cyclicity and trend? What do you learn about the series?**

```{r}
autoplot(myts)
```

The plot shows seasonality and a upward trend.

```{r}
ggseasonplot(myts) + ggtitle("Seasonal plot: retail sales")
```

The seasonal plot shows the there is a large jump in sales in December. The graph also shows that the sales decrease in February and increase in March.

```{r}
ggsubseriesplot(myts) + ggtitle("Seasonal subseries plot: retail sales")
```

The plot shows higher sales in December, February has low sales compared to other months. 

```{r, fig.height = 5}
gglagplot(myts, lags = 30)
```

Here the colours indicate the months. All other lags have positive corelation but the relationship is strongly positive at lags 12 and lag 24.

```{r}
ggAcf(myts, lag = 48) 
```

The slow decrease in the ACF as the lags increase is due to the trend, while the scalloped shape is due the seasonality. 

### Forecasting: Principles and Practices (Chapter 6 - Time serise Decomposition)

#### 6.2

**The plastics data set consists of the monthly sales (in thousands) of product A for a plastics manufacturer for five years.**

**$(a)$ Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?**

```{r}
autoplot(plastics)
```

The plot shows the seasonality and a upward trend. 

**$(b)$ Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.**

```{r, fig.height=5}
plastics %>% decompose(type="multiplicative") %>%
  autoplot() + xlab("Year") +
  ggtitle("Classical multiplicative decomposition
    of plastic sales")
```

**$(c)$ Do the results support the graphical interpretation from part a?**

Yes, the result support the graphical interpretation, above decomposition graph shows the trend, seasonal, and remainder. 

**$(d)$ Compute and plot the seasonally adjusted data.**

```{r}
dcomp <- plastics %>% decompose(type="multiplicative")
adjusted_seasonality <- plastics / dcomp$seasonal

autoplot(plastics, series = 'Data') +
    autolayer(adjusted_seasonality, series = 'Seasonality Adjusted') +
  labs(y = "Sales", x = "Year", 
       title = "Plastic sales", color = 'Serise') +
  scale_color_manual(name="Series", 
                        values = c("Data"="gray50", 
                                   "Seasonality Adjusted"="steelblue"))
```

**$(e)$ Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?**

```{r}
plastics2 <- plastics
plastics2[15] <- plastics2[15] + 500

dcomp2 <- plastics2 %>% decompose(type="multiplicative")
adjusted_seasonality2 <- plastics2 / dcomp2$seasonal

autoplot(plastics, series = 'Data') +
    autolayer(adjusted_seasonality2, series = 'Seasonality Adjusted') +
  labs(y = "Sales", x = "Year", 
       title = "Plastic sales", color = 'Serise') +
  scale_color_manual(name="Series", 
                        values = c("Data"="gray50", 
                                   "Seasonality Adjusted"="steelblue"))
```

The outlier has major impact on the plot. The sesonality adjusted line structure has changed.  

**$(f)$ Does it make any difference if the outlier is near the end rather than in the middle of the time series?**

```{r}
plastics3 <- plastics
plastics3[60] <- plastics3[60] + 500

dcomp3 <- plastics3 %>% decompose(type="multiplicative")
adjusted_seasonality3 <- plastics3 / dcomp3$seasonal

autoplot(plastics, series = 'Data') +
    autolayer(adjusted_seasonality3, series = 'Seasonality Adjusted') +
  labs(y = "Sales", x = "Year", 
       title = "Plastic sales") +
  scale_color_manual(name="Series", 
                        values = c("Data"="gray50", 
                                   "Seasonality Adjusted"="steelblue"))
  
```

We can see the outlier effect in the plot but at the end. The seasonality-adjusted line looks smooth, with not much variation in the line. 

